The extended light profiles of the Sculptor and Ursa Minor dwarf galaxies: Nature or tidal nurture?

### 2025-10-02 thesis/discussion.tex
discussion revisions

In the previous chapters, we have shown that (1) Sculptor and Ursa Minor
have unusually extended density profiles for MW classical satellites and
(2) recent tides are an unlikely explanation. Here, we first consider
the reliability and assumptions supporting our conclusions. We then
consider possible formation scenarios of extended density profiles,
including multi-epoch star formation, dwarf mergers and subhalo
encounters, and alternative theories of dwarf galaxy structure. We
finish with an outlook on prospects for disentangling various stellar
halo formation scenarios.

While tides have been a popular explanation for some features of dwarf
galaxies \citep[discussion above, and
e.g.,][]{tsujimoto+shigeyama2002, mayer+2001a}, our work, and many
others, have disfavoured this for Scl, UMi, and many other MW dwarfs.
For example, \citet{read+2006} suggest that the lack of observational
evidence for a rising velocity dispersion profile with radii indicates
most dwarfs are not tidally affected. Later, \citet{penarrubia+2009}
define and use the break radius, Eq.~\ref{eq:r_break}, to show that the
(then understood) orbits of most satellites are inconsistent with a
tidal feature. In addition, \citet{pace+erkal+li2022} use a criterion
based on the observed mean density and orbital pericentre (like the
Jacobi radius) to show most dwarfs are unlikely to be undergoing
significant tidal disruption. Most recently,
\citet{tchiorniy+genina2025} use a similar idealized framework to ours,
with a focus on the inner density, showing tides do not strongly affect
equilibrium assumptions for several classical dwarfs. The general tidal
evolution of our simulations is furthermore consistent with much of the
literature \citep[e.g.,][]{robles+bullock2021, EN2021}. Tides seem to
now play a subdominant role in the evolution of most Milky Way
satellites.

@@

For both Scl and UMi, these galaxies have been studied extensively in
both a theoretical and observational context. While many works
considered mass modelling, star formation histories, chemistry, and
other observational properties, we focus on the observations and models
concerning tides and dynamical evolution here.

However, some cosmological simulations have suggested that tidal effects
may be ubiquitous. \citep{riley+2024, shipp+2023} find that a majority
of satellites form substantial streams, and not due to resolution
effects \citep[see also][]{panithanpaisal+2021}. This result appears to
be in tension with ours and many previous works. Some reasons
cosmological and idealized simulations may give opposite results could
include resolution effects \citep[e.g.,][]{santos-santos+2025},
additional perturbations from dark matter substructure, tidal
``pre-processing'' by other infalling satellites, or the streams may be
below detection limits \citep[e.g.,][]{shipp+2023}. More work is needed
to understand why streams appear to form more readily in cosmological
simulations than is observed in the Milky Way or through idealized
simulations.

Closely related to our work, \citet{iorio+2019} applied idealized N-body
simulations to study tidal effects on Scl. They similarly found weak
tidal effects, even for a dark-matter-free model. However, their orbits
are less well-constrained and they did not consider the LMC.

\subsection{Sculptor}\label{sculptor}

Scl has long been speculated to be disturbed. \citet{innanen+papp1979}
found RR Lyrae candidate Scl members \citep[from][]{vanagt1978} out to
180' in an elongated distribution, speculating this to be tidal
disruption. Many later density profile determinations noted Scl's
elongation and apparent outer density excess similar to J+24's detection
\citep[but see also
\citet{coleman+dacosta+bland-hawthorn2005}]{eskridge1988, IH1995, walcher+2003, westfall+2006}.
Scl's ellipticity and extended density profile were often interpreted as
evidence of tidal debris or sometimes a dwarf galaxy halo.

In addition, Scl hosts at least two populations, as revealed through
photometry \citep{tolstoy+2004}, kinematics
\citep{battaglia+2008, tolstoy+2023, arroyo-polonio+2024}, dynamical
structure \citep{breddels+helmi2014}, age gradients \citep{deboer+2011},
and chemistry \citep{kirby+2009}. These two populations both have radii
smaller than the outer density excess. Scl challenges the idea of a
simple, single-component dwarf spheroidal population and hints at
episodic star formation or hierarchical assembly.

\subsection{Ursa Minor}\label{ursa-minor}

UMi has garnered claims of inner substructure such as stellar or
kinematic ``clumps''. Studies by \citet{olszewski+aaronson1985},
\citet{demers+1995}, \citet{IH1995}, \citet{kleyna+1998},
\citet{battinelli+demers1999}, and \citet{bellazzini+2002} note that
Ursa Minor appears to contain ``clumps'' along its major axis. One clump
has furthermore been shown to be kinematically distinct and colder
\citep[e.g.,][]{pace+2014}. If a star cluster, than the cluster should
disolve in \(\sim 3\) Gyrs provided the UMi's halo is not cored and did
not interact with dark (sub)subhalos---the survival of such substructure
depends on the nature of dark matter \citep{kleyna+2003, lora+2012}.
\citet{wilkinson+2004} additionally find a puzzling drop in the velocity
dispersion in the outskirts. The nature of any clumps or substructure
remains unclear, but \citet{munoz+2018}'s are unable to find signs of
any substructure with modern, deeper photometry.

In addition, several works have found supposed tidal features in UMi.
\citet{hargreaves+1994} first detected a velocity gradient in UMi,
suggestive of tidal disruption. Later, \citet{martinez-delgado+2001}
find that stars extend far beyond the ``tidal radius'' (from a King
profile fit) for Ursa Minor, in the direction of the galaxy's elongation
\citep[see corresponding simulations
by][]{gomez-flechoso+martinez-delgado2003}. \citet{palma+2003} further
showed evidence for S-shaped contours, and an extended population of
``extratidal'' stars. Our density profiles are consistent with these
works. However, strong evidence of tidal disruption has not yet been
found.

@@

As a tidal origin of Scl and UMi's extended density profiles is
disfavoured, we consider alternative processes which may form a dwarf
galaxy ``stellar halo''. Some recent works have also confirmed the
presence of extended, likely non-tidal, density profiles. For example,
\citet{chiti+2021}; \citet{chiti+2023} spectroscopically confirm members
out to \(~9 R_h\) in Tucana II. Given these stars are misaligned with
the orbit and the lack of a velocity gradient, tides seem to be an
unlikely explanation for Tucana II as well. The prevalence of stellar
halos around dwarf galaxies remains an active research topic.

Many dwarf galaxies also host multiple chemodynamical stellar
populations. Typically, older populations are extended, kinematically
hotter, and metal-poo , whereas the younger populations are more
compact, colder, and metal-rich. Examples include Carina
\citep[\citet{fabrizio+2016}, \citet{kordopatis+2016}]{battaglia+2012},
Fornax \citep[\citet{amorisco+evans2012},
\citet{delpino+aparicio+hidalgo2015}]{battaglia+2006}, Sextans
\citep{battaglia+2011, cicuendez+battaglia2018, roederer+2023}, and
Andromeda II
\citep{mcconnachie+arimoto+irwin2007, ho+2012, delpino+2017}. Evidence
of multiple-populations in dwarf galaxies suggests that effects like
galaxy mergers or episodic star formation shape the stellar structure of
dwarf galaxies.

We now explore possible origins of a ``stellar halo'' which may resemble
the extended density profiles in Scl and UMi.

\subsection{Internal processess}\label{internal-processess}

\textbf{Dynamical heating of old stars}. In a galaxy, older stars
generally have higher random velocities (i.e., kinematically ``hotter'')
than younger stars. In dwarf galaxies, several mechanism may heat
stellar components, including supernovae kicks in forming stars,
feedback-driven potential fluctuations
\citep{stinson+2009, maxwell+2012, el-badry+2016, mercado+2021}, or
heating by dark sub-subhalos \citep{penarrubia+2025}. Over time, these
processes naturally produce a more extended, older stellar population.

\textbf{Episodic star formation and feedback.} Dwarf galaxies are
thought to experience bursty star formation--i.e., consisting of several
episodes of intense star formation separated by periods of quiescence
\citep[e.g.,][]{salvadori+ferrara+schneider2008, valcke+derijcke+dejonghe2008, wheeler+2019, azartash-namin+2024}.
Stellar feedback in a dwarf galaxy's shallow potential well drives
oscillations in the star formation rate. Alternatively, re-ionization
may temporarily suspending star formation \citep{benitez-llambay+2015}.
In many simulations, shrinking gas reservoirs form
centrally-concentrated later generations of stars, naturally spawning
multiple populations \citep{kawata+2006, revaz+jablonka2018}. However,
star formation is also sensitive to a dwarf's environment.

\subsection{External processes}\label{external-processes}

\textbf{Induced star formation.} Star formation may quench and reignite
due to an external perturbation. Examples include tidal compression
\citep{mayer+2001a, dong+lin+murray2003}, gaseous filaments
\citep{genina+2019}, dark halos \citep{starkenburg+helmi+sales2016}, or
shocks with the MW corona \citep{wright+2019}. Induced burst would be
more stochastic than internal bursts, possibly explaining diversity in
dwarf galaxy structure.

\textbf{Major mergers.} When galaxies merge, they may leave signatures
such as population gradients and stellar halos. Classical dwarfs have a
\(\sim 10\%\) chance of undergoing a major merger since redshift \(z=1\)
\citep{deason+wetzel+garrison-kimmel2014}. Mergers may preferentially
disperse the lower-mass galaxy's stars, forming a halo and a metallicity
gradient \citep{benitez-llambay+2016}. Intermediate-mass mergers (with
mass-ratios of \(\sim\) 1:5) most effectively populate halos, balancing
stellar mass from larger galaxies with better dispersion of lower-mass
galaxies \citep{deason+2022}. Tuc II's properties are suggested to
originate from a similar merger
\citep{tarumi+yoshida+frebel2021, querci+2025}.

\textbf{Gas-rich mergers.} If a merger occurs between gas-rich galaxies,
triggered star formation may occur in the aftermath
\citep[e.g.,][]{genina+2019}. And II and Phoenix have steep metallicity
gradients and unusual prolate rotation, theorized to result from mergers
of disky dwarfs \citep{lokas+2014, fouquet+2017, cardona-barrero+2021}.

\textbf{Minor mergers / accretion}. Like how the Milky Way's halo is
believed to be built from many minor mergers, dwarf galaxies may form
halos through accretion of yet fainter dwarfs.
\citet{ricotti+polisensky+cleland2022} demonstrated that accretion of
ultra-faint dwarfs may create dwarf stellar halos. A stream detected
around And II further supports the occurrence of mergers among dwarfs
\citep{amorisco+evans+vandeven2014, roederer+2023}. In this scenario,
the halo chemistry would resemble a population of ultra-faint dwarfs.
The occurance of dwarf mergers would further constrain small-scale
galaxy formation.

\textbf{Tidal preprocessing}. Dwarf galaxies may have been
``preprocessed'' by a larger satellite like the LMC
\citep[e.g.,][]{santistevan+2023, riley+2024}. From the orbit
integrations above, it is possible that UMi was once an LMC satellite.
However, the prevalence of preprocessing remains uncertain. Like stellar
heating, preprocessing redistributes already-present stellar
populations.

\textbf{Tidal dwarf galaxies} are cluster-like objects which form in
gas-rich tidal streams created during the merger of two galaxies
\citep[e.g.,][]{mirabel+dottori+lutz1992, bournaud+duc2006}. Tidal
dwarfs may be more susceptible to tides, forming extended density
profiles and appearing to have dark matter
\citep{casas+2012, yang+2014, wang+2024a}. If Scl and UMi are tidal
dwarfs, a stronger velocity (dispersion) gradient and tidal tails should
be detectible.

\subsection{\texorpdfstring{Beyond \LCDM{}}{Beyond }}\label{beyond}

\textbf{Modified Newtonian Dynamics (MOND)}. MOND modifies gravity
instead of using dark matter to explain the rotation curves of galaxies.
In MOND, tides may more strongly affect dwarf galaxies owing to the lack
of dark matter mass and the stronger MW tidal field
\citep{mcgaugh+wolf2010, brada+milgrom2000}. A tidal origin of the
density excess is more likely in this case. If MOND is to recover the
observed velocity dispersions of satellites, then many more dwarfs may
be actively tidally disrupting as well
\citep{casas+2012, yang+2014, wang+2024a}.

\textbf{Self-interacting dark matter (SIDM).} SIDM significantly
complicates tidal evolution. SIDM halos in isolation are not static---by
transferring heat through collisions, these halos first undergo core
formation and then core collapse. Besides structural changes, SIDM adds
that pressure from the host DM halo can change the structure
(e.g.~aiding core collapse) or remove material from the inner galaxy
(analogous to ram pressure stripping) \citep[e.g.,][]{cartonzeng+2024}.
An SIDM halo may be more strongly affected by tides, but the velocity
dispersion makes a tidal disruption in SIDM still unlikely.

\textbf{Fuzzy dark matter} can also heat up stars owing to density
perturbations in the form of interference fringes
\citep[e.g.,][]{el-zant+2020, duttachowdhury+2023}. Similar to other
intrinsic heating methods, this model would likely affect all dwarfs
similarly, so we should detect similar halos around similar dwarfs.

\subsection{Disentangling the origin of a stellar
halo.}\label{disentangling-the-origin-of-a-stellar-halo.}

@@

\textbf{Chemistry}. Large samples of detailed chemical abundances have
been invaluable for understanding MW substructure (e.g.~for
\emph{Gaia}-Sausage Enceladus). Chemistry, particularly comparing the
inner and outer regions could test if the halo appears to originate from
a distinct system than the dwarf, differentiating many \emph{internal}
versus \emph{external} scenarios.

\textbf{Star formation histories}. Evidence of significant star
formation episodes or lack thereof may differentiate scenarios which
rely on strong star formation bursts (e.g.~\emph{episodic star formation
history, induced star formation, and gas-rich mergers.})

\textbf{Kinematics.} Particularly, for models relying on recent tidal
disruption, kinematic disequilibrium features should be visible. These
would appear as velocity gradients, increasing velocity dispersions,
outward-biased moving stars, or non-phase mixed structures
\citep[e.g,][]{kroupa1997, read+2006, sanchez-salcedo+hernandez2007}.

\textbf{Deep photometry} may find or rule out signs of dynamical
disequilibrium and tidal tails for tidally susceptible models
\emph{(e.g., MOND and tidal dwarfs.)}

Ongoing, upcoming, and future facilities will be essential for testing
these theories. For example, given the typical faint magnitudes of dSph
stars, chemistry is best accessible through large field-of-view
multi-object spectroscopic instruments on large telescopes (e.g., 4MOST
and extremely large telescopes \citet{skuladottir+2023}). 3D internal
kinematics of dwarfs may require a successor to \emph{Gaia}. However,
using JWST and HST enables precise proper motions for small regions
within dwarf galaxies \citep[e.g.,][]{vitral+2025}, which may be
sufficient to constrain vastly different internal dynamic structures and
ongoing tidal disruption. Finaly, Photometric surveys by Rubin
Observatory and Euclid will most-emminently probe yet fainter magnitudes
around dwarf galaxies and possibly find or constrain stellar halos and
their star formation histories. Altogether, new surveys will likely
uncover novel aspects about the inner workings and outer halos of dwarf
galaxies.

In addition to new observing facilities, the next generation of dwarf
simulations are beginning to answer questions and unravel processes in
the formation of dwarf galaxies. With improved resolution, realism, and
physics, the frequency and effects of various mechanisms discussed can
be better constrained.

@@

In this thesis, we have investigated the extended, outer density excess
of Sculptor and Ursa Minor and its possible tidal origin. We first
verified that Scl and UMi have unusually extended density profiles,
compared to other dwarf spheroidals. We show that the density profiles
are robust to alternate data criteria, implying that this ``density
excess'' is likely a real feature of each galaxy.

We then investigated if tides were a permissible explanation. By
modelling each galaxy based on cosmological initial conditions, we
showed that tides do not strongly affect either galaxy. The LMC changes
the orbital history of Scl and UMi, and tides become even weaker in a
combined LMC and MW potential. While UMi may form a stellar stream, the
stream is far fainter than is presently detectible. We conclude that
recent tides are unlikely to shape the observed stellar distributions of
Scl and UMi.

Finally, we consider alternative scenarios forming extended density
distribution. We review mechanisms ranging from episodic star formation
histories, to accretion, to departures from \LCDM{}. While a more
precise explanation awaits, upcoming and ongoing surveys will uncover
the stellar populations within dwarf galaxies in unprecedented detail.
Future work will the Milky Way, uncovering the cosmic origins of the
smallest structures forming building blocks of galaxies like ours.

>>>>

In the previous chapters, we have shown that Sculptor and Ursa Minor
have more extended density profiles compared to other MW classical
satellites, and that recent tides are an unlikely explanation for such
extended density profiles. We therefore suggest that Scl and UMi host a
``stellar halo''---an additional, extended population of stars. With our
data, other classical dwarfs lack evidence for similar stellar halos.
The absence of similar halos in other classical dwarfs suggests Scl and
UMi have followed a divergent evolutionary path.

In this chapter, we place our results in context with past work. We
review both work considering tidal effects on Scl and UMi and detections
of peculiarities. We then consider possible origins of stellar halos for
these galaxies. Scenarios include induced star formation forming
multiple populations, past mergers depositing stars into the outskirts,
and even departures from the standard dwarf galaxy structure. We later
discuss how future observations and simulations may distinguish halo
formation scenarios. Finally, we summarize the main conclusions of this
thesis.

@@

\subsection{The role of tides}\label{the-role-of-tides}

Tidal effects have been long invoked to explain features of dwarf
galaxies e.g. \ref{sec:scl_umi_obs_tides},
\citep{tsujimoto+shigeyama2002, mayer+2001a}. Yet, growing work suggests
tides may not be as commonplace as suspected. For example,
\citet{read+2006} suggest that the absence of rising radial velocity
dispersion profiles in dwarfs implies a lack of tidal disruption.
\citet{penarrubia+2009}, introducing the break radius as a diagnostic
(Eq.~\ref{eq:r_break}), showing that most satellite orbits were
inconsistent with observable tidal effects. Recently,
\citet{pace+erkal+li2022} use a Jacobi radius-based criterion
(Eq.~\ref{eq:r_jacobi}), reaching a similar conclusion. With the
exception of the Sagittarius dSph, these results suggest tides have mild
influence on the stellar components of classical dwarfs.

A few works have modelled Scl or UMi specifically, finding similar
conclusions. \citet{iorio+2019} applied idealized N-body simulations to
study tidal effects on Scl. They similarly found weak tidal effects,
even for a dark-matter-free model. Most recently,
\citet{tchiorniy+genina2025} also use idealized simulations tuned to
five classical dwarfs with a focus on the inner density, concluding that
tides do not strongly affect equilibrium assumptions or the stellar
component.

Our work extends these models, adding updated structural properties, a
broader range of orbital histories, and the influence of the LMC.
Despite these considerations, we still reach similar conclusions: tidal
effects do not shape Scl or UMi's stellar component. A non-tidal
explanation likely underlies the origin of these galaxies' density
excess.

\subsection{Peculiarities in the
galaxies}\label{peculiarities-in-the-galaxies}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/scl_umi_fe_h_gradient.pdf}}
\caption[Metallicity gradients in Scl and UMi]{Density profiles and
metallicity gradients in Scl (left) and UMi (right). In the top panel,
we plot the J+24 density profiles as function of elliptical radius in
black, and the two components of a double-exponential density fit as
curved, transparent lines. The vertical lines represent the transition
from the inner to the outer exponential profile. \textbf{Bottom}:
Metallicies as a function of elliptical radius for stars from
\citet{tolstoy+2023}; \citet{apogee}; and \citet{sestito+2023} (left)
and \citet{pace+2020}; \citet{apogee}; and \citet{sestito+2023b}
(right)}\label{fig:metallicity_gradients}
\end{figure}

Besides extended stellar density profiles, Scl and UMi display other
peculiarities which may hold clues to their formation. Scl and UMi both
hosts at least two chemodynamic populations, as revealed through their
photometric or metallicity-velocity structure
\citep{tolstoy+2004, battaglia+2008, pace+2020}. The inner, compact
population is younger, more metal-rich, and dynamically colder, whereas
the outer, extended population older, more metal-poor, and dynamically
hotter.

Could these internal stellar populations map to the observed density
excess? In Scl, both populations have scale radii likely smaller than
the half-light radius. But, the third, more tentatively detected
population in \citet{arroyo-polonio+2024} would reside within the
stellar halo. In Ursa Minor, the transition between the metal-rich and
metal-poor stellar components occurs at \(\sim 30\,\)arcmin
\citep{pace2020}, consummate with the beginning of the density excess.
This metal-poor population may thus be related to the stellar halo of
Ursa Minor. Fig:metallicity\_gradients shows the radial metallicity
distribution in both galaxies. Scl and UMi both show evidence of a
metallicity gradient, but the outskirts have few stars. Further
observations would be necessary to understand if the stellar halo
represents a chemically distinct population.

Ursa Minor has also shown evidence for possible inner substructure such
as stellar or kinematic ``clumps''
\citep[e.g.,][]{olszewski+aaronson1985, demers+1995, kleyna+1998, battinelli+demers1999, bellazzini+2002}.
While one clump was re-detected kinematically in \citet{pace+2014},
\citet{munoz+2018} find no evidence of substructure with modern
photometry. In any case, the survival of a cold clump depends on the
nature of dark matter, disrupting quickly in a \LCDM{} halo
\citep{kleyna+2003, lora+2012}.

Given their multiple stellar populations and possible inner
substructure, this hints that Scl and UMi's distinct structure today
arises from an unusual past. A mechanism forming a stellar halo could
also explain multiple populations.

@@

As we disfavour a tidal origin for Scl and UMi's extended density
profiles, we now consider alternative processes which may form a stellar
halo in these dwarfs.

\textbf{Episodic star formation.} Star formation may quench and
reignite, creating successive stellar generations with differing
distributions. External star formation triggers include tidal
compression \citep{mayer+2001a, dong+lin+murray2003}, collisions with
gaseous filaments \citep{genina+2019}, perturbations from dark halos
\citep{starkenburg+helmi+sales2016}, or shocks with the MW corona
\citep{wright+2019}. More common mechanisms, like feedback or
reionization-driven quenching, may also form multiple stellar
generations
\citep{kawata+2006, benitez-llambay+2015, revaz+jablonka2018}. However,
such processes would not explain why stellar halo formation may be
non-universal. If a stellar halo was formed in an older, induced burst
of star formation, then the star formation history should contain
evidence of a corresponding burst.

\textbf{Major mergers.} Dwarf galaxy mergers may be relatively common
\citep{deason+wetzel+garrison-kimmel2014}. After a merger, stars from
the lower-mass galaxy are preferentially dispersed, forming a stellar
halo and population gradient \citep{benitez-llambay+2016, deason+2022}.
If the galaxies contains gas, the merger can trigger new star formation,
forming a younger population of stars \citep[e.g.,][]{genina+2019}. A
few local dwarfs are suspected to have undergone a major merger,
including Tucana II, Andromeda II, and Phoenix
\citep{lokas+2014, fouquet+2017, tarumi+2021, cardona-barrero+2021, querci+2025}.
A galaxy having undergone a major merger should show evidence of at
least two or three populations from distinct origins.

\textbf{Minor mergers / accretion}. Just as the Milky Way's halo is
believed to be built from many minor mergers, dwarf galaxies may build
stellar halos composed of even fainter satellites
\citep{ricotti+polisensky+cleland2022}. Few examples are known, but a
stream around And II may be an instance of dwarf accretion
\citep{amorisco+evans+vandeven2014, roederer+2023}. In this scenario,
the stellar halo's would contain chemical signatures from several
distinct ultra-faint dwarf galaxies.

\textbf{Tidal preprocessing}. Some dwarf galaxies may have been tidally
``preprocessed'' by a massive satellite like the LMC
\citep[e.g.,][]{santistevan+2023, riley+2024}. Tidal preprocessing
redistributes already-present stellar populations without creating new
populations, and may mimic a stellar halo. Key evidence suggesting
preprocessing may include distant dwarf stars or stellar streams.

\textbf{Dark-matter free dwarfs.} Tidal dwarfs form in gas-rich tidal
streams created during the merger of two galaxies
\citep[e.g.,][]{mirabel+dottori+lutz1992, bournaud+duc2006}. Without a
dark matter halo, tidal dwarfs may be more susceptible to tides, forming
extended density profiles and inflating the velocity dispersion
\citep{casas+2012, yang+2014, wang+2024a}.

Similarly, in Modified Newtonian Dynamics (MOND), galaxies do not
contain dark matter. Instead the gravitational force law is altered to
explain rotation curves. Dwarfs would similarly experience stronger
tidal effects in MOND \citep{mcgaugh+wolf2010, brada+milgrom2000}.

In both cases, tides more plausibly produce extended density profiles.
However, recovering the observed velocity dispersions often requires
ongoing tidal disruption
\citetext{\citealp{mcgaugh+wolf2010}; \citealp[but see
also][]{sanchez-salcedo+hernandez2007}}. Current data do not show
evidence of such features (see Section~\ref{sec:extra_rv_models}), so
future data would need to uncover signs of disruption to support this
hypothesis.

\textbf{Dynamical heating.} Old stars in dwarf galaxies may be hotter
than younger stars due to processes including stellar feedback
\citep{stinson+2009, maxwell+2012, el-badry+2016, mercado+2021},
sub-subhalo interactions \citep{penarrubia+2025}, or even fuzzy dark
matter interference fringes
\citep[e.g.,][]{el-zant+2020, duttachowdhury+2023}. However, most of
these processes should operate similarly across dwarf galaxies---stellar
halos should be more common if these processes form substantial halos.

\section{Disentangling the origin of a stellar
halo}\label{disentangling-the-origin-of-a-stellar-halo}

@@

\emph{Precise chemistry}. Chemistry, particularly comparing the inner
and outer regions, could test if the halo appears to originate from a
distinct system than the dwarf. This would test if the halo is a
separate chemical population, may reveal the properties of any merged
galaxies, or show if the halo is consistent with originating from the
same galaxy.

\emph{Detailed kinematics}. For models relying on recent tidal
disruption, kinematic disequilibrium features should be visible. These
would appear as velocity gradients, increasing velocity dispersions,
outward-biased moving stars, or non-phase mixed structures
\citep[e.g,][]{kroupa1997, read+2006, sanchez-salcedo+hernandez2007}. In
addition, measurements of the ellipticity and anisotropy would help
verify or alter our understanding of the mass structure and if DM-free
models may be permissible.

\emph{Deep photometry} may find or rule out signs of dynamical
disequilibrium and tidal tails for tidally susceptible models
\emph{(e.g., pre-processing and DM-free dwarfs.)} In addition,
photometry will help constrain the prevalence and nature of extended
features in dwarf galaxies.

\emph{Star formation histories} can be derived through photometry or
chemistry. Evidence of significant star formation episodes or lack
thereof may differentiate scenarios which rely on strong star formation
bursts (e.g.~\emph{episodic star formation history, induced star
formation, and gas-rich mergers.})

Ongoing and future facilities will be essential for testing dwarf galaxy
halo formation. Given the typical faint magnitudes of dSph stars,
chemistry is best accessible through multi-object spectroscopic
instruments on large telescopes (e.g., 4MOST and extremely large
telescopes \citet{skuladottir+2023}). 3D internal kinematics of dwarfs
may require a successor to \emph{Gaia}. However, using JWST and HST
enables precise proper motions for small regions within dwarf galaxies
\citep[e.g.,][]{vitral+2025}, which may be sufficient to constrain
vastly different internal dynamic structures and ongoing tidal
disruption. Finally, ongoing photometric surveys by Rubin Observatory
\citep{ivezic+2019} and the Euclid mission
\citep{euclidcollaboration+2025} will probe even fainter features around
dwarf galaxies, possibly finding more halos, tidal streams, and
improving our understanding of their stellar populations. Altogether,
new surveys will likely uncover novel aspects about the inner workings
and outer halos of dwarf galaxies.

The next generation of dwarf simulations are beginning to answer
questions and unravel processes in the formation of dwarf galaxies.
Notable recent simulation projects include EDGE \citep{agertz+2020} and
LYRA \citep{gutcke+2021}, both of which can simulate the cosmological
formation of isolated dwarf galaxies while resolving individual
(massive) stars. By including more realism (e.g., black holes, magnetic
fields), refining models of poorly understood processes (feedback and
star formation), and improving numerical resolution, simulations will
continue to explore these processes and refine our understanding of
dwarf galaxy formation.

@@

In this thesis, we have investigated the extended, outer density excess
of Sculptor and Ursa Minor and its possible tidal origin. We show that
the density profiles are robust, implying that Scl and UMi contain a
true density excess relative to other classical dwarfs.

We then investigated if tides are a possible explanation. By modelling
each galaxy based on cosmological initial conditions, we showed that
tides do not strongly affect either galaxy. The LMC changes the orbital
history of Scl and UMi, but tides become even weaker in a combined LMC
and MW potential. We conclude that recent tides are unlikely to shape
the outer radial distribution of stars in Scl and UMi.

Finally, we discuss the implications of a non-tidal density excess in
Scl and UMi. We review mechanisms ranging from episodic star formation
histories, to accretion, to departures from \LCDM{}. While a more
precise explanation awaits, upcoming and ongoing surveys will uncover
the stellar populations within dwarf galaxies in unprecedented detail.
Future work will continue to uncover the cosmic origins of the smallest
structures forming building blocks of galaxies like ours.

### 2025-10-02 thesis/extra_density.tex
discussion revisions

In this Appendix, we describe the details of J+24's likelihoods. We then
test the reliability of the derived density profiles of the dwarfs,
comparing methodologies, samples, and to literature. In conclusion, the
density profiles from J+24 converge among methods out to a
background-limited radius, derived here.

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_literatre_profiles.pdf}}
\caption[Scl literature density profiles]{A comparison of the Scl
derived density profile and historical works.}
\end{figure}

@@

>>>>

In this section, we describe the details of J+24's likelihoods. We then
test the reliability of the derived density profiles for Scl and UMi,
comparing methodologies, samples, and to literature. Finally, we present
a comparison of J+24's samples to a non-parametric Bayesian-derived
density profile, finding similar results except in the very outer
regions. Past a ``limiting radius'', the density profiles become
susceptible to the likelihood specification and may select mostly
background stars. In conclusion, the density profiles from J+24 converge
among methods out to our limiting radii derived here.

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_literatre_profiles.pdf}}
\caption[Scl literature density profiles]{A comparison of the Scl
derived density profile and historical works. Unlike most profiles in
this thesis, this density profile is plotted with respect to the
semi-major elliptical radius (\(a = R_{\rm ell} / \sqrt{1-{\rm ell}}\)).
The solid black line is a 2D exponential with corresponding scale radius
to Scl's half-light radius, and residuals in the lower panel are
relative to this exponential.}
\end{figure}

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_literature_profiles.pdf}}
\caption[UMi density profiles in literatures]{Similar to}
\end{figure}

### 2025-10-02 thesis/extra_numerical_convergence.tex
discussion revisions

Here, we describe some convergence tests to ensure our methods and
results are minimally impacted by numerical limitations and assumptions.
See \citet{power+2003} for a detailed discussion of various assumptions
and parameters used in N-body simulations.

@@

where \(h_{\rm grav}\) is the softening length, and \(N_{200}\) is the
number of particles within \(R_{200}\). This choice balances integration
time and only compromises resolution in collisional regime.

For our isolation halo (\(M_s=2.7\), \(r_s=2.76\)) and with \(10^7\)
particles, this works out to be \(0.044\,{\rm kpc}\).We adpoted the
slightly smaller softening which was reduced by a factor of
\(\sqrt{10}\) which appears to improve agreement slightly in the
innermost regions.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/iso_converg_softening.png}}
\caption{Softening convergence}\label{fig:softening_convergence}
\end{figure}

\section{Time stepping and force
accuracy}\label{time-stepping-and-force-accuracy}

In general, we use adaptive timestepping and relative opening criteria
for gravitational force computations. To verify that these choices and
associated accuracy parameters minimally impact convergence or speed, we
show a few more isolation runs (using only 1e5 particles)

\begin{itemize}
\tightlist
\item
  constant timestep (\ldots), approximantly half of minimum timestep
  with adaptive timestepping
\item
  geometric opening, with \(\theta = 0.5\).
\item
  strict integration accuracy, (facc = \ldots.)
\end{itemize}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/iso_converg_methods.png}}
\caption{Isolation method convergence}\label{fig:methods_convergence}
\end{figure}

>>>>

Here, we describe the numerical specifics of our simulation and
analysis. We present convergence tests supporting our choices of
softening and additional parameters. We find alternate choices of
numerical parameters are unlikely to improve our convergence or result
in significantly different evolution.

@@

where \(h_{\rm grav}\) is the softening length, and \(N_{200}\) is the
number of particles within \(R_{200}\). This choice balances integration
time and only compromises resolution in collisional regime. For our
isolation halo (\(\rmax=6\,\kpc\), \(\vmax=31\,\kms\)) with \(10^7\)
particles, this works out to be \(0.044\,{\rm kpc}\), about
\(\sqrt{10}\) times larger than our adopted softening
(Eq.~\ref{eq:softening_length}).

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/iso_converg_softening.png}}
\caption[Softening convergence]{Similar to
Fig.~\ref{fig:numerical_convergence} except for simulations with
different softening lengths, \(h_{\rm grav}\). \(0.044\,\kpc\) is our
fiducial softening length for this halo. All simulations here were ran
in isolation of 5 Gyr with \(10^{6}\)
particles.}\label{fig:softening_convergence}
\end{figure}

Fig.~\ref{fig:softening_convergence} illustrates the influence of
softening length on isolation evolution. The simulation with
\(h=0.14\,\kpc\) deviates the most from the expected profile. Instead,
the final profiles for \(h=0.044\,\kpc\) and \(h=0.014\,\kpc\) are
nearly identical. Smaller values of softening than \(h=0.044\,\kpc\)
likely will not affect the evolution in the inner regions. As
computation time increases moderately with decreasing softening length,
\(h=0.044\,\kpc\) is an ideal choice balancing efficiency and accuracy.
Our simulations therefore adopt this choice (as in
Eq.~\ref{eq:softening_length}).

\section{Time-stepping and force
accuracy}\label{time-stepping-and-force-accuracy}

In general, we use adaptive time-stepping and relative opening criteria
for gravitational force computations. To verify that these choices and
associated accuracy parameters minimally impact convergence or speed, we
show a few more isolation runs (using only \(10^5\) particles).

We use in \gadget{} the relative tree opening criterion with the
accuracy parameter set to \(\alpha =0.005\) (so nodes are opened when
\(M\,l/r^3 < \alpha |a|\) for a distance from the node of mass \(M\) of
\(r\), \(l\) is the side length of the node, and \(a\) is the particle's
total acceleration), and adaptive time stepping with integration
accuracy set to \(\eta=0.01\) (particles must take time-steps smaller
than \(dt < \sqrt{2\,\eta\,h_{\rm grav} / a}\) for acceleration \(a\)).

Fig.~\ref{fig:methods_convergence} tests various additional simulation
parameters in isolation. We compare against an equivalent model ran
instead with \texttt{GADGET-2}, a model ran with constant timestep of
0.5Myr (\texttt{dt0.1} named after our dimensionless units), a
simulation with the time-step accuracy set to \(\eta=0.003\) (high
accuracy), and a simulation using a geometric opening criterian instead
of the force (geometric). In all cases, the final isolation velocity
profile is nearly the same after 5Gyr. The numerical details likely do
not limit our convergence (as marked by the arrow).

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/iso_converg_methods.png}}
\caption[Isolation method convergence]{A comparison against different
integration methods. See text for details on each
model.}\label{fig:methods_convergence}
\end{figure}
